---
output:
  html_document: default
  pdf_document: default
---
Homework 1 (Collective Exercises)
Amy Burkel
599B Data Visualization

```{r eval=FALSE}
#install libraries
install.packages("haven")
```

```{r 2 eval=FALSE}
#call libraries:
library(haven)
library(ggplot2)
```

```{r 3}
#call datasets:
link="https://github.com/cazvan/VisualAnalytics_2_tabularData/raw/master/Small%20Dataset-Caz/policylong.dta"
link2='https://github.com/cazvan/VisualAnalytics_2_tabularData/raw/master/Small%20Dataset-Caz/policywide.dta'
link3='https://github.com/cazvan/VisualAnalytics_2_tabularData/raw/master/Small%20Dataset-Caz/statedata.dta'
link4='https://dl.dropboxusercontent.com/s/xhs2utpj1wntmzt/BRFSSFinalSup.dta'
brfssdata=read_dta(link4)
policylong=read_dta(link)
policywide=read_dta(link2)
statedata=read_dta(link3)
```
1) Use ggplot to show a bar plot ordered by share size.
2) Complete adding the elements missing in the last plot. (title, source)
3) Make sure our box plot follows the same design approach and include all the elements as in the bar plot for nominal data. 
4) Make a histogram for the variable FREE LUNCH, and make sure it has all the right elements, and get rid of unnecessary elements.
5) Do some research and make a histogram and a density plot using ggplot for the variable we just used above.

###########################################################################################
1) Use ggplot to show a bar plot ordered by share size:

```{r}
library(ggplot2)

year=table(brfssdata$year)
yrd=sort(year)
yrd
```

```{r}
yrdgg=as.data.frame(yearord)
names(yrdgg)=c("Year","Responses")
yrdgg
```

```{r}
Yearplot= ggplot(data = yrdgg, 
             aes(x =Year,
                 y = Responses )) 

plot1 = Yearplot + geom_bar(fill ="gray",
                        stat = 'identity') # y is just what it is!
plot1
Yearplot= ggplot(data = tableFreq, 
             aes(reorder(x=Year,-Responses),
                 y = Responses)) 
             
plot1 = Yearplot + geom_bar(fill ="gray",
                        stat = 'identity') # y is just what it is!


```

###########################################################################################
2) Complete adding the elements missing in the last plot.

```{r}
base = ggplot(tableFreq, aes(x=locale,pct)) 
lolliplot1=base + geom_segment(aes(y = 0, 
                                   x = locale, 
                                   yend = pct, 
                                   xend = locale), color = "grey50") 
lolliplot1 + geom_point()

tableFreq[order(tableFreq$pct),]

tableFreqO=tableFreq[order(tableFreq$pct),]


base = ggplot(tableFreqO, aes(locale,pct)) 
lolliplot1=base + geom_segment(aes(y = 0, 
                                   x = locale, 
                                   yend = pct, 
                                   xend = locale), color = "gray") 
lolliplot2 = lolliplot1 + geom_point()
lolliplot2 + scale_x_discrete(limits=tableFreqO$locale) # key element

# new variable
tableFreqO$gap=tableFreqO$pct-25

# plot the new variable
base = ggplot(tableFreqO, aes(locale,gap)) 

lolliplot1=base + geom_segment(aes(y = 0, 
                                   x = locale, 
                                   yend = gap, 
                                   xend = locale), color = "gray") 
lolliplot2 = lolliplot1 + geom_point()
lolliplot2 + scale_x_discrete(limits=tableFreqO$locale) # key element

# a new column for color
tableFreqO$PositiveGap=ifelse(tableFreqO$gap>0,T,F)

# add new aesthetics 'color'
base = ggplot(tableFreqO, aes(locale,gap,
                              color=PositiveGap)) #change
lolliplot1=base + geom_segment(aes(y = 0, 
                                   x = locale, 
                                   yend = gap, 
                                   xend = locale), color = "gray") 
lolliplot2 = lolliplot1 + geom_point()
lolliplot2 + scale_x_discrete(limits=tableFreqO$locale) # key element

# a new column for color
tableFreqO$PositiveGap=ifelse(tableFreqO$gap>0,T,F)

base = ggplot(tableFreqO, aes(locale,gap,color=PositiveGap,
                              label = round(gap,3))) #  change
lolliplot1=base + geom_segment(aes(y = 0, 
                                   x = locale, 
                                   yend = gap, 
                                   xend = locale), color = "gray") 
lolliplot2=lolliplot1 + geom_point() 
lolliplot3= lolliplot2 + scale_x_discrete(limits=tableFreqO$locale) 
# annotating and moving the text on the horizontal
lolliplot3 + geom_text(nudge_x=0.3)
lolliplot3 + geom_text(nudge_x=0.3,show.legend = FALSE) 


title3= "A Graph with Labels"
source3='Data from Washington School Districts'
lolliplot4= lolliplot3 + labs(title=title3, x='Type of School', y= 'Percentage from 25% baseline',caption=source3)
lolliplot4
```


###########################################################################################
3) Make sure our box plot follows the same design approach and include all the elements as in the bar plot for nominal data.

```{r}
#### <span style="color:blue"> 2. Visualization for ordinal scales</span>
#For this section, we will use the variable that tells us the highest grade offered in a school. A simple exploration gives:
table(eduwa$High.Grade,exclude = 'nothing')
#Being a categorical variable, the default option is again the bar plot. So let's prepare the frequency table as a data frame:
frqTabO=as.data.frame(prop.table(table(eduwa$High.Grade)))
names(frqTabO)=c('grade','pct')
frqTabO
#Now, we can use ggplot:
base = ggplot(frqTabO,aes(x=grade,y=pct))
base + geom_bar(stat = 'identity') 
#The x-values in this variable have **order**. That is, there is an increasing level in the values. Whenever we have an ordering, besides *concentration* we can visualize **symmetry**: if there is bias towards lower or higher values. 
#Bar plots help you see concentration and symmetry, but we have an alternative way to clearly detect symmetry, via **boxplots**:
# boxplots do not use frequency tables
# as.numeric produces turns levels of the factor into numbers
box1 = ggplot(eduwa, aes(y=as.numeric(High.Grade))) 
box1 = box1 + geom_boxplot() + coord_flip() # to show it horizontally
box1
#You have symmetry when the distance of those whiskers to the  box is the same, and when the thick line is in the middle of the box. You can see that the values show a negative asymmetry (tail to the left).
#Box plots expect a numeric value as an input, but we have an ordered categorical, so we used the _as.numeric()_ function. However, that eliminated the levels we saw in the previous bar plot; we can put the levels back in our plot:
# the labels use the original ordinal levels
ordLabels= levels(eduwa$High.Grade)
box2 = box1 + scale_y_continuous(labels=ordLabels,breaks=1:15) 


box2


title4='Grade Percentage Distribution'
source4='Data from Washington School Districts'
box3 = box2 + labs(title=title4, y='Type of School', x=NULL, caption=source4)
box3
#Box plots have important statistical information. The beginning and the ending of the box indicates the first (q1) and the third quantile (q75); and the thicker line in the middle represents the median. All those values are clearly visible, but we can retrieve the values like this:
#get positions
# using 'ggplot_build'
pos_q1=     ggplot_build(box2)$data[[1]]$lower
pos_median= ggplot_build(box2)$data[[1]]$middle
pos_q3=     ggplot_build(box2)$data[[1]]$upper
# using
levels(eduwa$High.Grade)[c(pos_q1,pos_median,pos_q3)]
#From the information retrieved, we know:
#* 25% of the public Schools offer at most 5th GRADE. 
#* 50% of the public Schools offer at most 8th GRADE.
#* 75% of the public Schools offer at most 12th GRADE. Also, 25% of the schools offer at least 12th grade.
#We can find these results with a _detailed_ frequency table; that is, instead of using the command _table_ as we did before, we could try a more advanced function:
library(summarytools)
freq(eduwa$High.Grade,style = 'rmarkdown')

```


###########################################################################################
4) Make a histogram for the variable  FREE LUNCH, and make sure it has all the right elements, and get rid of unnecessary elements.

```{r}
summary(eduwa$Free.Lunch)

eduwa3=eduwa[complete.cases(eduwa$Free.Lunch),]
dataHist=hist(eduwa3$Free.Lunch) #saving info in dataHist

base= ggplot(eduwa3,aes(x = Free.Lunch))  
h1= base + geom_histogram(binwidth = 120,boundary=0) #changing width
h1= h1 + stat_bin(binwidth = 120, aes(label=..count..), 
                  geom = "text",boundary = 0,vjust=-0.1,)

title5='Free Lunch Program'
source5='Washington School District Data'
h2= h1 + labs(title=title5, caption=source5,x='Children ', y= 'Free Lunch')
h2

```


###########################################################################################
5) Do some research and make a histogram and a density plot using ggplot for the variable we just used above.

```{r}
mv=as.data.frame(meanValues)
names(mv)=c("STratio")
head(mv)
base=ggplot(mv,aes(x=STratio))

base+geom_histogram()
base+geom_density()
```













